Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-02-13 paper.Rnw
Mallar's comments

We conduct two separate Monte Carlo cross-validation experiments comparing \mb{}
and \multiatlas{} whole hippocampal segmentation using differing atlas and
template library sizes, and registation and label fusion methods. The first
experiment is a 10-fold validation (per parameter setting) over 60 subjects
taken from the Alzheimer's Disease Neuroimaging Database (ADNI), and the second
is a five-fold validation over 81 subjects having had a first episode of psychosis.
In both cases, automated segmentations are compared with manual segmentions
following the \prsnr{}-protocol. Using the best settings found from these
experiments, we segment \Sexpr{length(means.complete$RID)} images of the
\adnidataset{} dataset and compare these with segmentations from existing
automated methods: \fsl{}, \freesurfer{}, \maper{}, and \snt.  Finally, we
conduct a leave-one-out cross-validation (LOOCV) of hippocampal subfield
segmentation in standard 3T T1 images, using five high-resolution manually
segmented atlases \citep{Winterburn2013}.

\textbf{Results:} In the ADNI cross-validation, using 9 atlases \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) score of \Sexpr{mb.kappa.best} with
respect to manual whole hippocampus segmentations, and also exhibits
significantly lower variability in DSC scores than \multiatlas{} segmentation.
In the younger, psychosis dataset, \mb{} achieves a mean DSC score of
\Sexpr{fep.mb.kappa.best} and produces volumes which agree with manual
segmentation volumes better than those produced by the \freesurfer{} and \fsl{}
methods (mean difference in volume: $80mm^3$, $1600mm^3$, and $800mm^3$,
respectively). Similarily, in the \adnidataset{} dataset, \mb{} produces
hippocampal segmentations well correlated ($r > 0.85$) with \snt{} semi-automated
reference volumes within disease categories, and shows a conservative bias and a
mean difference in volume of $250 mm^3$ across the entire dataset, compared with
\freesurfer{} and \fsl{} which both overesimate volume differences by $2600
mm^3$ and $2800 mm^3$ on average, respectively. Finally, \mb{} segments the CA1,
CA4/DG and subiculum subfields with an DSC overlap scores of 0.56, 0.65, 0.58,
respectively.

\textbf{Conclusion:} We demonstrate that \mb{} produces accurate whole
hippocampal segmentations using at most only 9 atlases with various hippocampal
definitions, disease populations, and image acquisition types, as well we show
that accurate identification of the hippocampal subfields is possible.
\end{abstract}

@@

Finally, we have provided evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,Yushkevich2010,VanLeemput2009}. Table
\ref{tab:subfield-segmentation-kappa} compares segmentation accuracy from some
of these methods and \mb{}. The overlap DSC scores for \mb{} subfields are
notably lower but a direct comparison of overlap values must be done cautiously.
In the present work, our overlap scores are computed on $0.9mm$-isotropic voxel
resolution images, whereas \citet{Yushkevich2010} uses focal
$0.4\times0.5\times2.0mm$ voxel resolution images, and \citet{VanLeemput2009}
use supersampled $0.380mm$-isotropic voxel resolution images. The larger voxel
images we use necessarily entail a greater change in DSC for each incorrectly
labelled voxel. In addition, our automated segmentations are compared to manual
segmentations resampled from $0.3mm$-isotropic voxel labels; the resampling
process inevitably introduces noise which may lower overlap scores. Lastly, as
our method is aimed specifically at situations when manually produced atlases
are scarce, in our cross validation we are forced to use three rather than all
five of the Winterburn atlases (which, based on our findings with whole
hippocampal segmentation, would lead to improved overlap similarity).
Although having more atlases would be ideal in this context, these atlases are
incredibly time consuming to generate \citep{Winterburn2013}. Nevertheless, the
advantage of evaluating \mb{} on standard 3T T1-weighted resolution MR images
with a publically available atlas library is that our results reflect typical
usage scenarios of researchers and clinicians.

@@

This points to a larger issue of how to truly validate subfield segmentations,
both in high resolution images and in standard T1-weighted images. There are
several manual subfield segmentation methodologies, and they do not agree on
which regions can be differentiated, even on high-resolution scans. See Table
\ref{tab:subfield-segmentation-protocols} for a comparison of manual subfield
segmentation methodologies. A further complication is that different researchers
have differing operational definitions for the subfields and how they ought to
be parcellated. The disagreement in the community has led to an international
working group devoted to normalizing the ontology and segmentation rules for the
hippocampal subfields (\url{http://www.hippocampalsubfields.com/}). Both of
these disagreements suggest that direct comparison of automated methods using
``ground truth''-based overlap similarity metrics, such as Dice's Similarity
Coefficient, are not possible without carefully taking into account the
differences in underlying segmentation protocols and image characteristics.

>>>>

We conduct two separate Monte Carlo cross-validation experiments comparing \mb{}
and \multiatlas{} whole hippocampal segmentation using differing atlas and
template library sizes, and registation and label fusion methods. The first
experiment is a 10-fold validation (per parameter setting) over 60 subjects
taken from the Alzheimer's Disease Neuroimaging Database (ADNI), and the second
is a five-fold validation over 81 subjects having had a first episode of psychosis.
In both cases, automated segmentations are compared with manual segmentions
following the \prsnr{}-protocol. Using the best settings found from these
experiments, we segment \Sexpr{length(means.complete$RID)} images of the
\adnidataset{} dataset and compare these with segmentations from existing
automated methods: \fsl{}, \freesurfer{}, \maper{}, and \snt.  Finally, we
conduct a leave-one-out cross-validation (LOOCV) of hippocampal subfield
segmentation in standard 3T T1-weighted images, using five high-resolution
manually segmented atlases \citep{Winterburn2013}.

\textbf{Results:} In the ADNI cross-validation, using 9 atlases \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) score of \Sexpr{mb.kappa.best} with
respect to manual whole hippocampus segmentations, and also exhibits
significantly lower variability in DSC scores than \multiatlas{} segmentation.
In the younger, psychosis dataset, \mb{} achieves a mean DSC score of
\Sexpr{fep.mb.kappa.best} and produces volumes which agree with manual
segmentation volumes better than those produced by the \freesurfer{} and \fsl{}
methods (mean difference in volume: $80mm^3$, $1600mm^3$, and $800mm^3$,
respectively). Similarily, in the \adnidataset{} dataset, \mb{} produces
hippocampal segmentations well correlated ($r > 0.85$) with \snt{} semi-automated
reference volumes within disease categories, and shows a conservative bias and a
mean difference in volume of $250 mm^3$ across the entire dataset, compared with
\freesurfer{} and \fsl{} which both overesimate volume differences by $2600
mm^3$ and $2800 mm^3$ on average, respectively. Finally, \mb{} segments the CA1,
CA4/DG and subiculum subfields on standard 3T T1-weighted resolution images with
DSC overlap scores of 0.56, 0.65, and 0.58, respectively, relative to manual
segmentations.  %TODO: use Sexpr

\textbf{Conclusion:} We demonstrate that \mb{} produces accurate whole
hippocampal segmentations using only 9 atlases, or fewer, with various
hippocampal definitions, disease populations, and image acquisition types.
Additionally, we show that \mb{} identifies hippocampal subfields in standard 3T
T1-weighted images with overlap scores comparable to competing methods.
\end{abstract}

@@

Finally, we have provided evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,Yushkevich2010,VanLeemput2009}. Table
\ref{tab:subfield-segmentation-kappa} compares segmentation accuracy from some
of these methods and \mb{}. The overlap DSC scores for \mb{} subfields are
notably lower but a direct comparison of overlap values must be done cautiously.
In the present work, our overlap scores are computed on $0.9mm$-isotropic voxel
resolution images, whereas \citet{Yushkevich2010} uses focal
$0.4\times0.5\times2.0mm$ voxel resolution images, and \citet{VanLeemput2009}
use supersampled $0.380mm$-isotropic voxel resolution images. The larger voxel
images we use necessarily entail a greater change in DSC for each incorrectly
labelled voxel. In addition, our automated segmentations are compared to manual
segmentations resampled from $0.3mm$-isotropic voxel labels; the resampling
process inevitably introduces noise which may lower overlap scores. Lastly, as
our method is aimed specifically at situations when manually produced atlases
are scarce, in our cross validation we are forced to use three rather than all
five of the Winterburn atlases (which, based on our findings with whole
hippocampal segmentation, would have resulted in improved overlap similarity).
Although having more atlases would be ideal in this context, these atlases are
very time consuming to generate \citep{Winterburn2013}. Nevertheless, the
advantage of evaluating \mb{} on standard 3T T1-weighted resolution MR images
with a publically available atlas library is that our results reflect typical
usage scenarios of researchers and clinicians.

@@

This points to a larger issue of how to truly validate subfield segmentations,
both in high resolution images and in standard T1-weighted images. There are
several manual subfield segmentation methodologies, and they do not agree on
which regions can be differentiated, even on high-resolution scans. See Table
\ref{tab:subfield-segmentation-protocols} for a comparison of MRI-based manual
subfield segmentation methodologies. A further complication is that different
researchers have differing operational definitions for the subfields and how
they ought to be parcellated. The disagreement in the community has led to an
international working group devoted to normalizing the ontology and segmentation
rules for the hippocampal subfields
(\url{http://www.hippocampalsubfields.com/}).  In addition, there have been
recent advances from the Yushkevich group to revise their MRI subfield
segmentation protocol based on anatomy discerned from serial histological
acquisitions \citep{Adler2014}.  The definitional and operational disagreements
suggest that direct comparison across automated methods using ``ground
truth''-based overlap similarity metrics, such as Dice's Similarity Coefficient,
are not possible without carefully taking into account the differences in
underlying segmentation protocols and image characteristics.
